Pollution, agricultural productivity, and development: Evidence from coal in plants in India^[All scripts used in this paper are publicly available on my GitHub page: github.com/JoshMerfeld/pollution_development]

### 2023-01-16 draft/draft.Rmd
updating draft

---
title: "Pollution, agricultural productivity, and development: Evidence from coal in plants in India^[All scripts used in this paper are publicly available on my GitHub page: github.com/JoshMerfeld/pollution_development]"
author:
  - Joshua D. Merfeld^[KDI School of Public Policy and Management and IZA; merfeld@kdis.ac.kr]
date: "`r Sys.Date()`"
abstract: |
  | \noindent The rapid increase in population in India from the 1990s through the 2010s led to the rapid construction of new coal power plants throughout the country. Using this roll-out combined with randdom changes in wind direction, I document negative effects of pollution from coal power plants on agricultural productivity throughout the country. Results suggest this decrease may be driven by a decline in labor allocation of the oldest half of my sample, though I cannot rule out direct effects of pollution on land productivity. These impacts spill over into aggregate economic growth, as well; nightlight growth is slower following years of higher exposure to pollution.
  |
  | \textbf{\textit{Keywords}}: pollution, productivity, agriculture, labor, India
  | \textbf{\textit{JEL Codes}}: H40, I15, J22, O13, Q52, Q53

@@

# Tables
plantresultstable <- readRDS("pollution_development/draft/tables/plantresultstable.rds")
pollutiontable <- readRDS("pollution_development/draft/tables/pollutiontable.rds")
yieldtable <- readRDS("pollution_development/draft/tables/yieldtable.rds")
yieldtabletwo <- readRDS("pollution_development/draft/tables/yieldtabletwo.rds")
yieldtablehet <- readRDS("pollution_development/draft/tables/yieldtablehet.rds")
pmeffects <- readRDS("pollution_development/draft/tables/yieldtablePM.rds")
labortable <- readRDS("pollution_development/draft/tables/labortable.rds")
labortablemonsoon <- readRDS("pollution_development/draft/tables/labortablemonsoon.rds")
labortablewinter <- readRDS("pollution_development/draft/tables/labortablewinter.rds")
labortableold <- readRDS("pollution_development/draft/tables/labortableold.rds")
ntltable <- readRDS("pollution_development/draft/tables/ntltable.rds")
yieldtableleads <- readRDS("pollution_development/draft/tables/yieldtableleads.rds")
yieldtablepostplant <- readRDS("pollution_development/draft/tables/yieldtablepostplant.rds")
laborplot <- readRDS("pollution_development/draft/tables/laborplot.rds")

@@

In this paper, I study the effects of emissions from coal power plants on agricultural productivity, labor allocation, and economic growth in India. Previous work has suggested that there may be harmful direct effects of air pollution [@heck1982assessment;@marshall1997hidden] and water pollution [@reddy2006impact] on agricultural productivity, including work showing that gold mines -- which release large amounts of pollutants into the surrounding environment -- lead to lower agricultural productivity in Ghana [@aragon2016polluting]. This paper, however, tackles the broader case of coal plants, which are a leading source of power throughout the world. I show that pollution from these plants can lead to decreased agricultural productivity in the case of India.

I combine the roll out of coal plants in India with plausibly exogenous changes in wind direction -- in a similar spirit to [@deryugina2019mortality] -- to measure the effects of pollution from coal plants on key development outcomes. I first locate the location of coal plants in India by year. Using village-level shapefiles, I identify all villages that are located within 30km of a coal plant at any timn during the sample period, which comprises the 1990s to the 2010s, with differences depending on data availability. I then download daily wind data and create an identifier for whether a village is located downwind from a coal plant on any given day, aggregating these daily values to different levels depending on the analysis.

@@

The first set of data lists the location of coal plant across the globe. This data comes from Global Energy Monitor.^[https://globalenergymonitor.org/projects/global-coal-plant-tracker/] and lists all units generating at least 30 megawatts of electricity. The data on coal plants includes the year of opening (and, if applicable, the year of retirement), the GPS (latitude/longitude) location of each plant, and the amount of power produced by the units at each plant. For this paper, I do not use the information on the capacity of the plant.

@@

Figure \@ref(fig:plants) presents the location of coal plants in India for two specific years: 1990 and 2010. There is a clear increase in the prevalence of coal plants across the country over the two decades. Additionally, the overall capacity from coal plants increased from `r paste0(reportDecOne(sum(plants1990_capacity$capacity)))` gigawatts to `r paste0(reportDecOne(sum(plants2010_capacity$capacity)))` gigawats, an increase of `r paste0(reportPercent((sum(plants2010_capacity$capacity) - sum(plants1990_capacity$capacity))/sum(plants1990_capacity$capacity)))` percent in just 20 years.

The second dataset includes agricultural productivity for both the monsoon and winter seasons, from 2002 to 2018. To match other data, I only use the data up to 2013. This data comes from @gangopadhyay2022new and estimates land productivity (i.e. yield, in tons per hectare) for the major crops in India. They define the monsoon season as June to October and the winter season as November to March. I keep these definitions when matching across data below. This data is also publicly available from Nature's data-sharing website.^[https://springernature.figshare.com]

@@

To measure exposure to pollution from coal plants, I first locate all village centroids located within 30km from a coal plant in a given year. I choose 30km due to previous research on the effects of (air) pollution [@aragon2016polluting;@bencko1977health;@li2014health]. I then calculate the direction from coal plants to all village centroids within that 30km radius. To define exposure, I then pull daily wind direction data from the National Center for Atmospheric Research.^[https://climatedataguide.ucar.edu/] For each day, I document whether the wind is blowing towards each village centroid.^[I define "towards" as within five degrees to help take into account that I am using village centroids, using the x and y components of wind speed.] I then temporally aggregate this daily data depending on the temporal definition of the corresponding outcome. For example, agricultural productivity is defined across five months (e.g. the monsoon season is from June to October) so I count the total days a given village is exposed to wind during those five months.

@@

To help understand some of the mechanisms driving the relationship between exposure to emissions and agricultural productivity, I also include household survey data. Unfortunately, I am not aware of any publicly available survey data that covers multiple years and has identifiers below the district level. For example, the ARIS-REDS data^[https://www.ncaer.org/project/additional-rural-incomes-survey-aris-rural-economic-demographic-survey-reds] includes GPS identifiers for one year, but that only allows GPS matching for a total of two years: 1999 and 2007/2008. The DHS, on the other hand, only has district-level identifiers (except for the newest two waves) and is also not a labor survey. Instead, I use the National Sample Survey (NSS) from the Ministry of Statistics and Programme Implementation.

@@

The second column keeps the 1991 census data but changes the outcome variable to a dummy for whether there *will be* a coal plant constructed before 2001. This column restricts the sample to only villages that do not already have a coal plant in 1991. Population is no longer a significant predictor while the coefficients on literacy and the presence of a paved road have flipped.

@@

I first present results for agricultural productivity. The outcome of all the regressions in this section is the log of agricultural land productivity, which is defined as tons per hectare. The first set of results are in Table \@ref(tab:yieldtable). The unit of analysis is the year season -- from 2002 to 2013 -- the geographic fixed effect is the village-season, and the temporal fixed effect is the year. The exposure variable is defined for the entire season. Since @gangopadhyay2022new define both the monsoon (kharif) season and the winter (rabi) season as five months long, I do as well. This implies that the exposure variable ranges from zero to slightly more than 150.

@@

@@

If you believe that wind direction is exogenous and, conditional on the fixed effects and rainfall, only affects agricultural productivity through pollution -- which seems reasonable since it is hard to imagine what other variables it might affect -- then it is a valid instrument for 
pollution. While I opt to present the reduced form results for the rest of this paper, it is nonetheless an interesting exercise to also use wind direction as an instrument for pollution, which I measure using estimated $\mathrm{PM_{2.5}}$ from @hammer2020global.

The IV results appear in Table \@ref(tab:yieldtabletwo), with the second-stage results in the top panel and the first-stage results in the bottom panel. The first-stage results make clear how strong of an instrument wind is; also note that the coefficients are quite different from those in Table \@ref(tab:pollutiontable) because the unit of analysis in Table \@ref(tab:yieldtabletwo) is the season, while in Table \@ref(tab:pollutiontable) the unit of analysis is the month.

The second-stage results indicate that $\mathrm{PM_{2.5}}$ is a very strong predictor of agricultural productivity. The smallest coefficient -- that for the monsoon season in column (4) -- and the mean absolute deviation for $\mathrm{PM_{2.5}}$ point to a change in agricultural productivity of around 23 percent.

@@

The first two columns split the sample based on average exposure; villages with maximum exposure above the median are included in column (1) while those below the median are in column (2). Interestingly, the largest effects are seen on villages in the lower half of the distribution. Apparently, smaller maximum effects lead to larger overall effects of pollution exposure on agricultural productivity. In other words, it is not the places where wind is most likely to blow from a coal plant that experience the largest decrease in productivity in response to an additional day of wind-driven pollution.

@@

In the final column I add to the regression an interaction between wind and initial agricultural productivity (defined in 2002, the first year observed in the productivity data). The linear term drops out of the regression since it is invariant within villages, leaving just the interaction term. The negative coefficient on the interaction term of -0.001 indicates that the effect of wind is more negative in areas where agricultural productivity is higher. Note that agricultural productivity is defined in logs, so this can be interpreted as a percentage effect, not a level effect. Apparently, wind-driven pollution is particularly detrimental to agricultural productivity in the most productive places.

>>>>

---
title: "Pollution, agricultural productivity, and development: Evidence from coal in plants in India^[All scripts used in this paper are publicly available on my GitHub page: github.com/JoshMerfeld/pollution_development]"
author:
  - Joshua D. Merfeld^[KDI School of Public Policy and Management and IZA; merfeld@kdis.ac.kr]
date: "`r Sys.Date()`"
abstract: |
  | \noindent The rapid increase in population in India from the 1990s through the 2010s led to the rapid construction of new coal power plants throughout the country. Using this roll-out combined with random changes in wind direction, I document negative effects of pollution from coal power plants on agricultural productivity throughout the country. Results suggest this decrease may be driven by a decline in labor allocation of the oldest half of my sample, though I cannot rule out direct effects of pollution on land productivity. These impacts spill over into aggregate economic growth, as well; nightlight growth is slower following years of higher exposure to pollution.
  |
  | \textbf{\textit{Keywords}}: pollution, productivity, agriculture, labor, India
  | \textbf{\textit{JEL Codes}}: H40, I15, J22, O13, Q52, Q53

@@

# Tables
plantresultstable <- readRDS("pollution_development/draft/tables/plantresultstable.rds")
pollutiontable <- readRDS("pollution_development/draft/tables/pollutiontable.rds")
yieldtable <- readRDS("pollution_development/draft/tables/yieldtable.rds")
yieldtablepm <- readRDS("pollution_development/draft/tables/yieldtablepm.rds")
yieldtabletwo <- readRDS("pollution_development/draft/tables/yieldtabletwo.rds")
yieldtablehet <- readRDS("pollution_development/draft/tables/yieldtablehet.rds")
pmeffects <- readRDS("pollution_development/draft/tables/yieldtablePM.rds")
labortable <- readRDS("pollution_development/draft/tables/labortable.rds")
labortablemonsoon <- readRDS("pollution_development/draft/tables/labortablemonsoon.rds")
labortablewinter <- readRDS("pollution_development/draft/tables/labortablewinter.rds")
labortableold <- readRDS("pollution_development/draft/tables/labortableold.rds")
ntltable <- readRDS("pollution_development/draft/tables/ntltable.rds")
yieldtableleads <- readRDS("pollution_development/draft/tables/yieldtableleads.rds")
yieldtablepostplant <- readRDS("pollution_development/draft/tables/yieldtablepostplant.rds")
laborplot <- readRDS("pollution_development/draft/tables/laborplot.rds")

@@

In this paper, I study the effects of emissions from coal power plants on agricultural productivity, labor allocation, and economic growth in India. Previous work has suggested that there may be harmful direct effects of air pollution [@heck1982assessment;@marshall1997hidden] and water pollution [@reddy2006impact] on agricultural productivity, including work showing that gold mines -- which release large amounts of pollutants into the surrounding environment -- lead to lower agricultural productivity in Ghana [@aragon2016polluting]. This paper, however, tackles the broader case of coal plants, which are a leading source of power throughout the world. I show that air pollution from these plants can lead to decreased agricultural productivity.

I combine the roll out of coal plants in India with plausibly exogenous changes in wind direction -- in a similar spirit to @deryugina2019mortality -- to measure the effects of pollution from coal plants on key development outcomes. I first locate the location of coal plants in India by year. Using a village-level shapefile, I identify all villages that are located within 30km of a coal plant at any time during the sample period, which comprises the 1990s to the 2010s, with differences depending on data availability. I then download daily wind data and create an identifier for whether a village is located downwind from a coal plant on any given day, aggregating these daily values to different levels depending on the analysis.

@@

The first set of data lists the location of coal plant across the globe. This data comes from Global Energy Monitor^[https://globalenergymonitor.org/projects/global-coal-plant-tracker/] and lists all units generating at least 30 megawatts of electricity. The data on coal plants includes the year of opening (and, if applicable, the year of retirement), the GPS (latitude/longitude) location of each plant, and the amount of power produced by the units at each plant. For this paper, I do not use the information on the capacity of the plant.

@@

Figure \@ref(fig:plants) presents the location of coal plants in India for two specific years: 1990 and 2010. There is a clear increase in the prevalence of coal plants across the country over the two decades. Additionally, the overall capacity from coal plants increased from `r paste0(reportDecOne(sum(plants1990_capacity$capacity)))` gigawatts to `r paste0(reportDecOne(sum(plants2010_capacity$capacity)))` gigawats, an increase of `r paste0(reportPercent((sum(plants2010_capacity$capacity) - sum(plants1990_capacity$capacity))/sum(plants1990_capacity$capacity)))` percent in just 20 years. This increase is driven both by the rollout of new plants as well as the construction of new units at existing plants, the latter of which is not shown in Figure \@ref(fig:plants).

The second dataset includes agricultural productivity for both the monsoon and winter seasons, from 2002 to 2018. To match other data, I only use the data up to 2013. This data comes from @gangopadhyay2022new and estimates land productivity (i.e. yield, in tons per hectare) for the major crops in India. They define the monsoon season as June to October and the winter season as November to March. I keep these definitions when matching across data below. This data is also publicly available from Nature's data-sharing website.^[https://springernature.figshare.com] Since the main regressions below use village fixed effects, any differences in cropping patterns across areas should not bias estimates on changes, as long as these patterns do not change markedly in response to pollution.

@@

To measure exposure to pollution from coal plants, I first locate all village centroids located within 30km from a coal plant in a given year. I choose 30km due to previous research on the effects of (air) pollution [@aragon2016polluting;@bencko1977health;@li2014health]. I then calculate the direction from coal plants to all village centroids within that 30km radius. To define exposure, I then pull daily wind direction data from the National Center for Atmospheric Research.^[https://climatedataguide.ucar.edu/] For each day, I document whether the wind is blowing towards each village centroid.^[I define "towards" as within five degrees to help take into account that I am using village centroids, using the x and y components of wind speed.] I then temporally aggregate this daily data depending on the temporal definition of the corresponding outcome. For example, agricultural productivity is defined across five months (e.g. the monsoon season is from June to October) so I count the total days a given village is exposed to wind blowing from _any_ coal plant within 30km during those five months.

@@

To help understand some of the mechanisms driving the relationship between exposure to emissions and agricultural productivity, I also include household survey data. Unfortunately, I am not aware of any publicly available survey data that covers multiple years and has identifiers below the district level. For example, the ARIS-REDS data^[https://www.ncaer.org/project/additional-rural-incomes-survey-aris-rural-economic-demographic-survey-reds] includes GPS identifiers for one year, but that only allows GPS matching for a total of two years: 1999 and 2007/2008. The DHS, on the other hand, only has state-level identifiers (except for the newest two waves) and is also not a labor survey. Instead, I use the National Sample Survey (NSS) from the Ministry of Statistics and Programme Implementation.

@@

The second column keeps the 1991 census data but changes the outcome variable to a dummy for whether there *will be* a coal plant constructed within 30km before 2001. This column restricts the sample to only villages that do not already have a coal plant within 30km in 1991. Population is no longer a significant predictor while the coefficients on literacy and the presence of a paved road have flipped.

@@

I first present results for agricultural productivity. The outcome of all the regressions in this section is the log of agricultural land productivity, which is defined as tons per hectare. The first set of results are in Table \@ref(tab:yieldtable). The unit of analysis is the vilalge-year-season -- from 2002 to 2013 -- the geographic fixed effect is the village-season, and the temporal fixed effect is the year. The exposure variable is defined for the entire season. Since @gangopadhyay2022new define both the monsoon (kharif) season and the winter (rabi) season as five months long, I do as well, using the same five months as the authors. This implies that the exposure variable ranges from zero to slightly more than 150.

@@

```{r yieldtablepm, include = TRUE, echo = FALSE, message = FALSE, warning = FALSE}

# Table
colnames(yieldtablepm) <- c("(1)", "(2)", "(3)", "(4)", "(5)")
kable(
      yieldtablepm, caption = "Particulate matter (PM 2.5) and agricultural productivity",
      align = "ccccc", booktabs = TRUE, linesep = ""
      ) %>%
  footnote(
           "Note: Standard errors are in parentheses and are clustered at the village level. * p<0.10 ** p<0.05 *** p<0.01", 
           general_title = "",
           threeparttable = TRUE,
           footnote_as_chunk = TRUE
           ) %>%
  column_spec(1, width = "3cm") %>%
  column_spec(c(2:6),width = "2cm") %>%
  row_spec(5, bold = TRUE) %>%
  row_spec(9, hline_after = TRUE) %>%
  add_header_above(c(" ", "all" = 3, "monsoon" = 1, "winter" = 1)) %>%
  kable_classic_2()

Of course, it is not wind, per se, that we are interested in. Instead, we are actually interested in better understanding the effects of exposure to pollution on agricultural productivity. Table \@ref(tab:yieldtablepm) presents the same regressions as Table \@ref(fig:yieldtable), with a key difference: instead of wind exposure, it uses average concentration of $\mathrm{PM_{2.5}}$ -- from @hammer2020global -- during the season as the key explanatory variable. I rescale particulate matter by dividing by one thousand to help with presentation of the coefficients. Several interesting patterns emerge. In the first column we see no relationship between particulate matter and agricultural productivity, while there is actually a positive relationship in columns two, four, and five. There is a small, negative relationship between pollution and productivity in column three.

Recall, however, the results from above: coal plants were placed in places with higher average agricultural productivity. This means that the simple regressions in \@ref(tab:yieldtablepm) might actually be biased upwards: although pollution might negatively affect productivity, pollution is higher in more productive areas due to the endogenous placement of coal plants in the study areas. To recover the actual effect of pollution, we need an instrument. If you believe that wind direction is exogenous and, conditional on the fixed effects and rainfall, only affects agricultural productivity through pollution -- which seems reasonable since it is hard to imagine what other variables it might affect -- then it is a valid instrument for pollution.

I present the IV results in Table \@ref(tab:yieldtabletwo), with the second-stage results in the top panel and the first-stage results in the bottom panel. The first-stage results make clear how strong of an instrument wind is; the t-statistic in column one, for example, is almost 48 while it is still around 11 in column three, meeting even the higher bar for the first-stage F-test argued for by @lee2022valid. Also note that the coefficients are quite different from those in Table \@ref(tab:pollutiontable) because the unit of analysis in Table \@ref(tab:yieldtabletwo) is the season, while in Table \@ref(tab:pollutiontable) the unit of analysis is the month.

The second-stage results first indicate that $\mathrm{PM_{2.5}}$ is a very strong predictor of agricultural productivity. The smallest coefficient -- that for the monsoon season in column (4) -- and the mean absolute deviation for $\mathrm{PM_{2.5}}$ point to a change in agricultural productivity of around 23 percent, which is much higher than the wind results documented above. This makes sense, of course, since the IV is essentially scaling up the effects of wind. The comparison of the results to those in the previous table also confirm the concerns around endogenous placement of coal plants. The coefficients in Table \@ref(tab:yieldtabletwo) are always more negative than those in Table \@ref(tab:yieldtablepm), and sometimes substantially so. In other words, the simple regression of agricultural productivity on pollution results in coefficients on pollution that are upwards biased, since coal plants are placed in more productive areas. The IV strategy corrects this bias and leads to a robust, negative relationship between pollution and agricultural productivity.

@@

The first two columns split the sample based on maximum exposure; villages with maximum exposure above the median are included in column (1) while those below the median are in column (2). Interestingly, the largest effects are seen on villages in the lower half of the distribution. Apparently, smaller maximum effects lead to larger overall effects of pollution exposure on agricultural productivity. In other words, it is not the places where wind is most likely to blow from a coal plant that experience the largest decrease in productivity in response to an additional day of wind-driven pollution.

@@

@@

In the final column I add to the regression an interaction between wind and initial agricultural productivity (defined in 2002, the first year observed in the productivity data). The linear term drops out of the regression since it is invariant within villages, leaving just the interaction term. The negative coefficient on the interaction term of -0.001 indicates that the effect of pollution is more negative in areas where agricultural productivity is higher. Note that agricultural productivity is defined in logs, so this can be interpreted as a percentage effect, not a level effect. Apparently, wind-driven pollution is particularly detrimental to agricultural productivity in the most productive places.
